---
title: "s443 group project"
author: "Jolin"
date: "2026-03-26"
output: html_document
---

-   **load and clean data set**
    -   All data starts from Jan 1997 to Dec 2025
-    Load libraries

```{r}
library(readr)
library(dplyr)
library(tidyr)
```

-   read [employment by industry]{.underline} data in r (unit: 1000 people)

```{r}
employment_data <- read.csv("employment1.csv", header = FALSE, check.names = FALSE)

industry_names_em <- employment_data[-1, 1]
value_mat_em <- employment_data[-1, -1]

value_mat_em[] <- lapply(value_mat_em, function(x) as.numeric(gsub(",", "", x)))
value_mat_em <- as.matrix(value_mat_em)

rownames(value_mat_em) <- industry_names_em
```

-   transform employment data of each industry into ts object

```{r}
# person in thousands

total_ts <- ts(value_mat_em["Total, all industries", ], start = c(1997, 1), frequency = 12)

goods_ts <- ts(value_mat_em["Goods-producing sector", ], start = c(1997, 1), frequency = 12)

utilities_ts <- ts(value_mat_em["Utilities", ], start = c(1997, 1), frequency = 12)

construction_ts <- ts(value_mat_em["Construction", ], start = c(1997, 1), frequency = 12)

manufacturing_ts <- ts(value_mat_em["Manufacturing", ], start = c(1997, 1), frequency = 12)

durables_ts <- ts(value_mat_em["Durables", ], start = c(1997, 1), frequency = 12)

nondurables_ts <- ts(value_mat_em["Non-durables", ], start = c(1997, 1), frequency = 12)

services_ts <- ts(value_mat_em["Services-producing sector", ], start = c(1997, 1), frequency = 12)

wholesale_ts <- ts(value_mat_em["Wholesale trade", ], start = c(1997, 1), frequency = 12)

retail_ts <- ts(value_mat_em["Retail trade", ], start = c(1997, 1), frequency = 12)

transport_ts <- ts(value_mat_em["Transportation and warehousing", ], start = c(1997, 1), frequency = 12)

finance_ts <- ts(value_mat_em["Finance and insurance", ], start = c(1997, 1), frequency = 12)

real_estate_ts <- ts(value_mat_em["Real estate and rental and leasing", ], start = c(1997, 1), frequency = 12)

professional_ts <- ts(value_mat_em["Professional, scientific and technical services", ], start = c(1997, 1), frequency = 12)

educational_ts <- ts(value_mat_em["Educational services", ], start = c(1997, 1), frequency = 12)

health_ts <- ts(value_mat_em["Health care and social assistance", ], start = c(1997, 1), frequency = 12)

info_ts <- ts(value_mat_em["Information, culture and recreation", ], start = c(1997, 1), frequency = 12)

accommodation_ts <- ts(value_mat_em["Accommodation and food services", ], start = c(1997, 1), frequency = 12)

other_ts <- ts(value_mat_em["Other services (except public administration)", ], start = c(1997, 1), frequency = 12)

public_ts <- ts(value_mat_em["Public administration", ], start = c(1997, 1), frequency = 12)

## no matching gdp
business_ts <- ts(value_mat_em["Business, building and other support services", ], start = c(1997, 1), frequency = 12)

agriculture_ts <- ts(value_mat_em["Agriculture", ], start = c(1997, 1), frequency = 12)

forestry_ts <- ts(value_mat_em["Forestry, fishing, mining, quarrying, oil and gas", ], start = c(1997, 1), frequency = 12)

forestry2_ts <- ts(value_mat_em["Forestry and logging and support activities for forestry", ], start = c(1997, 1), frequency = 12)

fishing_ts <- ts(value_mat_em["Fishing, hunting and trapping", ], start = c(1997, 1), frequency = 12)

mining_ts <- ts(value_mat_em["Mining, quarrying, and oil and gas extraction", ], start = c(1997, 1), frequency = 12)
##

```

-   read [unemployment rate]{.underline} data (unit: 1000 people or percentage)

```{r}
unemployment_data <- read.csv("unemployment_rate.csv", header = FALSE, check.names = FALSE)

characteristic_names <- unemployment_data[-1, 1]

value_mat_un <- unemployment_data[-1, -1]

value_mat_un[] <- lapply(value_mat_un, function(x) as.numeric(gsub(",", "", x)))

value_mat_un <- as.matrix(value_mat_un)

rownames(value_mat_un) <- characteristic_names
```

-   transform unemployment rate data of each industry into ts object

```{r}

# person in thousands
population_ts <- ts(value_mat_un["Population", ], start = c(1997, 1), frequency = 12)

labour_force_ts <- ts(value_mat_un["Labour force", ], start = c(1997, 1), frequency = 12)

employment_ts <- ts(value_mat_un["Employment", ], start = c(1997, 1), frequency = 12)

full_time_employment_ts <- ts(value_mat_un["Full-time employment", ], start = c(1997, 1), frequency = 12)

part_time_employment_ts <- ts(value_mat_un["Part-time employment", ], start = c(1997, 1), frequency = 12)

unemployment_ts <- ts(value_mat_un["Unemployment", ], start = c(1997, 1), frequency = 12)

# rate --> percentage
unemployment_rate_ts <- ts(value_mat_un["Unemployment rate", ], start = c(1997, 1), frequency = 12)

# participation rate = unemployment rate + employment rate
participation_rate_ts <- ts(value_mat_un["Participation rate", ], start = c(1997, 1), frequency = 12)

employment_rate_ts <- ts(value_mat_un["Employment rate", ], start = c(1997, 1), frequency = 12)

```

-   read [GDP by industry]{.underline} data (unit: 1,000,000 cad)

```{r}
gdp_data <- read.csv("gdpindustry.csv", header = FALSE, check.names = FALSE)

industry_names_gdp <- trimws(as.character(gdp_data[-1, 1]))
value_mat_gdp <- gdp_data[-1, -1]

value_mat_gdp[] <- lapply(value_mat_gdp, function(x) as.numeric(gsub(",", "", x)))
value_mat_gdp <- as.matrix(value_mat_gdp)

rownames(value_mat_gdp) <- industry_names_gdp
```

-   transform GDP by industry industry into ts object

```{r}
total_gdp_ts <- ts(value_mat_gdp["All industries", ], start = c(1997, 1), frequency = 12)

goods_gdp_ts <- ts(value_mat_gdp["Goods-producing industries", ], start = c(1997, 1), frequency = 12)

services_gdp_ts <- ts(value_mat_gdp["Services-producing industries", ], start = c(1997, 1), frequency = 12)

nondurables_gdp_ts <- ts(value_mat_gdp["Non-durable manufacturing industries", ], start = c(1997, 1), frequency = 12)

durable_gdp_ts <- ts(value_mat_gdp["Durable manufacturing industries", ], start = c(1997, 1), frequency = 12)

utilities_gdp_ts <- ts(value_mat_gdp["Utilities", ], start = c(1997, 1), frequency = 12)

construction_gdp_ts <- ts(value_mat_gdp["Construction", ], start = c(1997, 1), frequency = 12)

manufacturing_gdp_ts <- ts(value_mat_gdp["Manufacturing", ], start = c(1997, 1), frequency = 12)

wholesale_gdp_ts <- ts(value_mat_gdp["Wholesale trade", ], start = c(1997, 1), frequency = 12)

retail_gdp_ts <- ts(value_mat_gdp["Retail trade", ], start = c(1997, 1), frequency = 12)

transportation_gdp_ts <- ts(value_mat_gdp["Transportation and warehousing", ], start = c(1997, 1), frequency = 12)

info_gdp_ts <- ts(value_mat_gdp["Information and cultural industries", ], start = c(1997, 1), frequency = 12)

finance_gdp_ts <- ts(value_mat_gdp["Finance and insurance", ], start = c(1997, 1), frequency = 12)

real_estate_gdp_ts <- ts(value_mat_gdp["Real estate and rental and leasing", ], start = c(1997, 1), frequency = 12)

professional_gdp_ts <- ts(value_mat_gdp["Professional, scientific and technical services", ], start = c(1997, 1), frequency = 12)

educational_gdp_ts <- ts(value_mat_gdp["Educational services", ], start = c(1997, 1), frequency = 12)

health_care_gdp_ts <- ts(value_mat_gdp["Health care and social assistance", ], start = c(1997, 1), frequency = 12)

accommodation_gdp_ts <- ts(value_mat_gdp["Accommodation and food services", ], start = c(1997, 1), frequency = 12)

other_gdp_ts <- ts(value_mat_gdp["Other services (except public administration)", ], start = c(1997, 1), frequency = 12)

public_gdp_ts <- ts(value_mat_gdp["Public administration", ], start = c(1997, 1), frequency = 12)


## no matching employment
industrial_production_ts <- ts(value_mat_gdp["Industrial production", ], start = c(1997, 1), frequency = 12)

energy_sector_ts <- ts(value_mat_gdp["Energy sector", ], start = c(1997, 1), frequency = 12)

agriculture_forestry_fishing_hunting_ts <- ts(value_mat_gdp["Agriculture, forestry, fishing and hunting", ], start = c(1997, 1), frequency = 12)

mining_quarrying_oil_gas_ts <- ts(value_mat_gdp["Mining, quarrying, and oil and gas extraction", ], start = c(1997, 1), frequency = 12)

administrative_support_waste_ts <- ts(value_mat_gdp["Administrative and support, waste management and remediation services", ], start = c(1997, 1), frequency = 12)

arts_entertainment_recreation_ts <- ts(value_mat_gdp["Arts, entertainment and recreation", ], start = c(1997, 1), frequency = 12)
##
```
